---
output:
  md_document:
    variant: gfm # markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  dev = "png",
  dpi = 150,
  fig.asp = 0.8,
  fig.width = 5,
  out.width = "60%",
  fig.align = "center"
)
```


## bsitar

The **bsitar** package provides an interface for Bayesian implementation of
Super Imposition by Translation and Rotation (SITAR) growth model. The SITAR is
a shape-invariant nonlinear mixed effect model that fits a natural cubic spline
mean curve and aligns individual-specific growth curves to the underlying
mean curve via a set of random effects: the size, timing and intensity. The
**bsitar** package package is a front-end to the R package **brms** which uses 
the **Stan** program (see https://mc-stan.org/) to performing full Bayesian
inference for a range of regression specifications including hierarchical
multivariate modeling.



## Installation

You can install the development version of bsitar from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Sandhu-SS/bsitar")
```

## Example

This is a basic example which shows you how to fit the model:

```{r example, echo = TRUE, eval = FALSE}
library(bsitar)
library(magrittr)
```


```{r data, echo = TRUE, eval = FALSE}
data(heights, package = "sitar")
data <- heights
rm(heights)

data <- data %>% 
  dplyr::select(id, age, height) %>% 
  tidyr::drop_na(height) %>% 
  dplyr::rename(y = height) %>% 
  dplyr::arrange(id, age) %>% 
  dplyr::mutate(id = as.factor(id) ) 
```


```{r fit1, echo = TRUE, eval = FALSE}
fit1 <- bsitar(age, y, id, data = data, df = 4, chains = 2, iter = 1000)
```

